Supplementary Materials

### 2023-07-02 manuscript/manuscript.tex
supp sim info

\begin{align}
  %V(x) &= U s - 2 {R}(x) s & \text{\emph{fitness variance equation}} \\
  V_a &= (U - 2 R) s_j & \text{\emph{fitness variance equation}} \\
  Q^2(m_i, s_j) &= \frac{2}{(1-Z)(2-(2-M)Z)} & \text{\emph{linkage inflation factor}} \\
  Z &= 1 - \frac{Us_j}{U-{M}} & \text{\emph{variance decay rate}}
\end{align}
%
and $U = m_i L$ and $M = r_\text{BP} L$ are the total mutation and
recombination rates in the segment. A detailed derivation of these equations
can be found in Supplementary Section \ref{suppsec:theory}. The recombination
rate in a segment is determined by a user-supplied recombination map.

@@

We use a binomial likelihood with probability $\bar{\pi}(b; \Psi) = \pi_0
\bar{B}(b; \mu, \mathbf{W})$, where bars indicate averages over some bin width.
With multiple features classes and a map $k(g)$ between segments $g$ to their
class, the reduction map is,

\begin{align}
    \log\left(B(x; W, s_j) \right) = - \frac{1}{2} \sum_{g \in G} \sum_{i=1}^{n_s} V_g(\mu W_{j, k(g)}, s_j) Q_g^2(m_i, s_j, r_{x, g}).
\end{align}

Here, $\mu$ is the mutation rate and $W_{j,k(g)}$ is a DFE entry in the
$\mathbf{W}$ is the $n_s \times K$ DFE matrix over the selection grid and
features. Since each column of the DFE $\mathbf{W}$ must sum to one, numeric
optimization was done in unconstrained $n_s-1$ dimensional softmax space. Our
binomial likelihood modeled the number of different pairwise comparisons
observed per window given the predicted pairwise diversity and total number of
comparisons implied per window (see Supplementary Materials Sections
\ref{supp:data-summary} and \ref{supp:likelihood}).

We tested multiple different optimization routines, finding that BOBYQA
outperformed alternatives \parencite{Powell_undated-ec}. We inspected and
confirmed convergence with diagnostic plots finding stable optima across 10,000
random starts (see Supplementary Materials Section \ref{suppsec:optima}). We
assessed model fit using out-sample predictive error, calculated by leaving out
a whole chromosome during fitting and predicting its diversity. To calculate
uncertainty, we used a block jackknife approach in 10 Mbp windows
(Supplementary Section \ref{supp:jackknife}).

@@

We use two types of forward simulations in this paper; both use SLiM version
4.0.1 \parencite{Haller2023-uk,Haller2019-vu}. First, we simulated a 100 kbp
segment in a population of $N=1000$ diploids across grids of mutation and
selection parameters to assess the accuracy of the SC16 theory. Each simulation
recorded the number of fixations and the fitness variance per generation, and
at the end $10N$ generations, the ancestral recombination graph was recorded so
downstream Python scripts could calculate diversity. Each of these segment
simulations was replicated $10,000$ times and averaged to compare against
theoretic expectations.

Second, we conducted forward simulations of negative selection on whole human
chromosomes. These simulations used the HapMap recombination map
\parencite{International_HapMap_Consortium2007-nu} and treated coding,
PhastCons, and UTR regions as an approximate set of conserved regions. We used
these whole-chromosome simulations for three purposes. First, we simulated
negative selection on chromosome 10 to confirm our method to estimate classic B
and new B' maps matched the average empirical reduction map across a grid of
mutation and selection parameters. Second, we evaluated our composite
likelihood method by simulating negative selection on the first five human
chromosomes, across grids of mutation and selection parameters. Multiple
independently-evolved chromosomes were combined into a synthetic genome in
order to have enough data for our procedure. Third, we conducted additional
simulations like these, but varying the dominance coefficient and expanding the
population by a factor of 1.004 each generation, starting $9.3$ generations in
to mimic the human expansion out-of-Africa (XXX). More details about these
simulations can be found in Supplementary Section \ref{supp:simulation}.

>>>>

\begin{align}
  %V(x) &= U s - 2 {R}(x) s & \text{\emph{fitness variance equation}} \\
  V_a &= (U - 2 R) s_j & \text{\emph{fitness variance equation}} \\
  Q^2(m_i, s_j) &= \frac{2}{(1-Z)(2-(2-M)Z)} & \text{\emph{linkage inflation factor}} \\
  Z &= 1 - \frac{Us_j}{U-{M}} & \text{\emph{variance decay rate}}
\end{align}
%
and $U = m_i L$ and $M = r_\text{BP} L$ are the total mutation and
recombination rates in the segment. A detailed derivation of these equations
can be found in Supplementary Section \ref{supp:theory}. The recombination
rate in a segment is determined by a user-supplied recombination map.

@@

Our binomial likelihood models the number of different pairwise comparisons
observed per window given the total number of pairwise comparisons. The
probability is given by $\bar{\pi}(b; \Psi) = \pi_0 \bar{B}(b; \mu,
\mathbf{W})$, where bars indicate averages over some bin width, and the
reduction is

\begin{align}
    \log\left(B(x; \mu, \mathbf{W}) \right) = - \frac{1}{2} \sum_{g \in G} \sum_{j=1}^{n_s} V_g(\mu W_{j, k(g)}, s_j) Q_g^2(m_i, s_j, r_{x, g}).
\end{align}
%
Here, $\mu$ is the mutation rate and $W_{j,k(g)}$ is a DFE entry in the
$\mathbf{W}$ is the $n_s \times K$ DFE matrix over the selection grid and
features. See Supplementary Materials Sections \ref{supp:data-summary} and
\ref{supp:likelihood} for more details.

Our method uses two tricks to improve optimization over the mutation and DFE
parameters. First, we find $B(x; \mu, \mathbf{W})$ is exponential over columns
of $\mu \mathbf{W}$, which allows for optimization over this smooth function
rather than the grid. Second, we use softmax to convert constrained
optimization over the DFE columns (which must sum to one) to unconstrained. We
tested multiple different optimization routines, finding that BOBYQA
outperformed alternatives \parencite{Powell2009-jm,Johnson2007-tl}. We
inspected and confirmed convergence with diagnostic plots finding stable optima
across 10,000 random starts (see Supplementary Materials Section
\ref{supp:optima}). We assessed model fit using out-sample predictive error,
calculated by leaving out a whole chromosome during fitting and predicting its
diversity. To calculate uncertainty, we used a block jackknife approach in 10
Mbp windows (Supplementary Section \ref{supp:jackknife}).

@@

We conducted forward simulations of negative selection on whole human
chromosomes to validate our method at two stages. First, we simulated negative
selection on chromosome 10 using a realistic recombination map and putatively
conserved features to confirm that our classic B and new B' maps matched the
average simulation reduction map across mutation and selection parameters.
Second, we evaluated our composite likelihood method by simulating negative
selection on the first five human chromosomes, across grids of fixed mutation
and selection parameters. We then combined these into a synthetic genome, and
overlaid mutations on the ARG. Then, we ran our likelihood methods on the
resulting allele count data to assess model accuracy. We ran additional
synthetic genome simulations like these to evaluate the impact of two model
violations: recessivity of deleterious mutations and expanding populations. For
the latter, after $9.3$ generations, we grew the population by factor of 1.004
each generation to mimic the human expansion out-of-Africa
\parencite{Gutenkunst2009-pg}. We did not simulate population bottlenecks since
our analyses showed little difference between bottlenecked out-of-Africa
samples (CEU and CHB) and YRI samples. More details about these and the segment
simulations shown in Figure \ref{fig:figure-1}A-C can be found Supplementary
Section \ref{supp:simulation}.

### 2023-07-02 manuscript/supplementary.tex
supp sim info

\subsection{Derivation of Approximate Drift $R^2$}

@@

\section{Simulations}

@@

\subsection{Chromosome 10 Simulations}

XXX
, using the HapMap recombination map (XXX) and in coding regions, PhastCons
regions, and UTRs as an approximate set of putatively conserved segments

We calculate the mean squared error (MSE) and relative error between
whole-chromosome simulations and the predicted B and B' maps by averaging
branch diversity in 10 kilobase windows (tskit, XXX) and dividing by the
neutral expectation of $4N$. The theoretic lower bound of the MSE is due the
noise in estimating the expected reduction map from simulations. For $n=100$
replicates, the sampling noise is minimal relative to the evolutionary noise in
the coalescence process \parencite{Tajima1983-gu}. We use this as the theoretic
lower bound assuming the maximum coalescence noise at $B=1$, $\var(\bar{B})
\approx \nicefrac{2}{9n}$.

>>>>

\subsection{Derivation of Approximate Coalescence $R^2$}
\label{supp:r2-coal}

@@

\section{Simulations}
\label{supp:simulation}

All of our forward simulations used SLiM version 4.0.1
\parencite{Haller2019-vu,Haller2023-uk} and all Eidos scripts are available in
the \path{slim_sims/} directory in the GitHub repository
(\url{https://github.com/vsbuffalo/bprime}). These Eidos scripts take command
line arguments for the parameters. Our simulation pipelines use YAML
configuration files that specify the parameter grids and other settings, which
are read and dispatched across a cluster using Snakemake
\parencite{Koster2012-iv}.

@@

To confirm the SC16 theory, we simulated a 100 kbp segment in a population of
$N=1000$ diploids across grids of mutation and selection parameters. The Eidos
script for these simulations can be found at
\path{slim_sims/region/region.slim} and the analysis notebook at
\path{notebooks/region_simulations.ipynb} in the GitHub repository
(\url{https://github.com/vsbuffalo/bprime}). We simulated over grids of
mutation rates ($\mu \in \{10^{-9.5}, 10^{-9}, 10^{-8.5}, 10^{-8}\}$) and
selection coefficients ($s \in \{10^{-1}, 10^{-1.5}, 10^{-2}, \ldots,
10^{-4.5}\}$). Selection was additive within a site and multiplicative across
sites, and recombination was fixed at $10^{-8}$ per basepair.

We replicated $10,000$ simulations for each parameter combination to average
and compare against theory. Each simulation recorded the number of fixations
and the fitness variance per generation. At the end $10N$ generations, the
ancestral recombination graph was recorded. Then, downstream Python scripts
calculate branch-length diversity using \texttt{tskit}
\parencite{Kelleher2018-qu}, and we estimate the reduction factor in a window
centered in the region with $B = \nicefrac{\pi}{4N}$ (since branch length
diversity implies $\mu=1$).

\subsection{Whole-chromosome Simulations}

Our whole-chromosome simulations use realistic human recombination maps and
putatively conserved segments. These annotation differ slightly from our
primary model fits, since these simulations were written before the method was
finalized. We use the HapMap recombination map
\parencite{International_HapMap_Consortium2007-nu}, PhastCons regions, and UTRs
as an approximate set of putatively conserved segments.

Our first set of simulations were used to validate that our methods were
producing accurate classic B and B' maps. We replicated $100$ whole-chromosome
simulations over grids of mutation rates ($\mu \in \{10^{-10}, 10^{-9.5},
\ldots, 10^{-8}, 10^{-7.5}\}$) and selection coefficients ($s \in \{10^{-1},
10^{-1.5}, 10^{-2}, \ldots, 10^{-6}\}$). For each parameter combination, we
calculate mean branch diversity in 10 kbp windows across the 100 replicates to
create the average simulation reduction map.

We then compare this average simulation map to our method's theoretic B' maps
(see Figure \ref{fig:figure-1} D and E). We calculate the MSE per 10 kbp window
to estimate the total rate across the chromosome. We can assess the accuracy
relative to a theoretic lower bound of the MSE if the noise were due to
coalescence noise. We approximate this by noting that the MSE can be decomposed
into a squared bias term and a variance term. If our methods are unbiased in
estimating the true reduction at a position, the MSE is just the stochastic
variance due to the $n=100$ replicates we simulate and the coalescence noise in
each one. \textcite{Tajima1983-gu} showed that variance in diversity due to
coalescence noise is much larger than sampling noise, and is given by
$\var(\pi) \approx \nicefrac{2\theta^2}{9}$ in an infinite sample. One can
derive an analogous equation for the variance of average simulation $\bar{B}$
across $n$ replicate simulations as, $\var(\bar{B}) \approx
\nicefrac{2B^2}{9n}$. We note that this assumes no recombination but this
approximation is fine over short scales (such as the 10 kbp scale used here).
